---
title: "Introduction to R"
subtitle: "Grand Rounds 08/16/2019"
tutorial:
  version: 0.2
output: 
  ioslides_presentation
runtime: shiny_prerendered
---

<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
slides > slide:not(.nobackground):after {
  content: '';
}
</style>

```{js, echo=FALSE}
$(function() {
   var editor;
   $('.ace_editor').each(function( index ) {
     editor = ace.edit(this);
     editor.setFontSize("24px");
   });
})
```

```{r setup, include=FALSE}
library(learnr)
library(RTutorialUSC)
library(dplyr)
library(readr)
library(haven)
library(ggplot2)
library(shiny)
data("dbts_df")
knitr::opts_chunk$set(echo = FALSE)
```

# Reading Data Into R

## Reading Data: Finding Files in R

To load data, first tell R where the file is on your computer. 

But to do that, we have to know where we are...

- `getwd()` returns current directory (i.e. folder) that R is looking at
- `list.files("folder location")` returns all files in "folder location"
- `setwd()` change folder R is looking at
- `file.exists("folder location")` checks if a file exists in "folder location"

```{r file-functions, exercise=TRUE, exercise.lines=5}
getwd()
setwd("C:/Users/gmweaver/Desktop/usc_project/")
list.files()
list.files("data")
```

## Reading Data: Loading CSV Files into R

`read.csv()` reads CSV files into R. 

In addition to the CSV file location, we may specify the following parameters:

- `header`: Indicates if first row of CSV has variable names
- `stringsAsFactors`: Whether columns with non-numeric data should be treated as character or factor variables (to be discussed later)

```{r load-csv, exercise=TRUE, exercise.lines=5}
setwd("c:/users/gmweaver/desktop/usc_project")
diabetes <- read.csv(file = "data/diabetes.csv")
```

## Reading Data: Loading CSV Files into R

What happened when you ran the command?

```{r load-csv2, eval=FALSE, echo=TRUE}
diabetes <- read.csv(file = "data/diabetes.csv")
```

1) `read.csv()` read a file called `diabetes.csv` into R
    - R functions accept arguments (i.e. `file=`, `header=`)
        - Arguments are either required or have default (modifiable) values
        - `file=` is a required argument
        - `header=` has a default value of `TRUE`, so we didn't need to specify

2) The data was assigned, `<-`, to a variable called `diabetes`
    - Use `<-` to assign values to variables
    - Use `=` to set argument values in functions
    - The `diabetes` variable is an R object can now be used to access, transform, and analyze the data

## An aside: R Documentation

Q: How did we know what arguments to use in `read.csv()`?

A: By running the command `?read.csv`, RStudio returns a help file...

- **Description**: high-level overview of what the function does

- **Usage**: how to use the function and what arguments it accepts

```{r docs, eval=FALSE, echo=TRUE}
read.csv(file, header = TRUE, sep = ",", quote = "\"",
         dec = ".", fill = TRUE, comment.char = "", ...)
```

- **Arguments**: more detailed information about each argument

- **Details**: additional description of the function and any information deemed important for the user

- **Value**: describes what the function returns

- **Examples**: self-containted snippets of R code to test using the function

## Reading Data: Loading Other File Types into R

Data comes in many shapes and sizes. Luckily, a number of R packages are available to handle the most common file formats.

But first, what is an **R package**?

- As R became more popular, users needed a way to share code
- R packages solved this problem by providing a standardized process to develop and distribute R code

Use `install.packages("package name")` to install an R package

- `library(package_name)` loads the functions into the R session
- Note: Run `library(package_name)` every time you start a new R session

## Reading Data: Loading Other File Types into R

R Package Example: Reading CSV files using functions from other packages

- `readr`: Part of the R Tidyverse
- `data.table`: Improved efficiency when manipulating larger data sets
- `readxl`: Functions to read Excel (.xls, .xlsx) files

As an example, we can install the `readr` package to read the diabetes CSV:

```{r load-csv3, eval=FALSE, echo=TRUE}
install.packages("readr")
library(readr)
```

```{r load-csv4, exercise=TRUE, exercise.lines=5}
setwd("c:/users/gmweaver/desktop/usc_project")
diabetes <- read_csv(file = "data/diabetes.csv")
```

## Reading Data: Loading Other File Types into R

The `haven` package can read/write SAS and STATA data sets!

Reading STATA `.dta` files only requires a small modification to the previous code

```{r load-stata, eval=FALSE, echo=TRUE}
install.packages("haven")
library(haven)
diabetes <- read_stata(file = "data/diabetes.dta")
```

```{r load-stata2, exercise=TRUE, exercise.lines=5}
setwd("c:/users/gmweaver/desktop/usc_project")
diabetes <- read_csv(file = "data/diabetes.csv")
write_dta(diabetes, path = "data/diabetes.dta")
diabetes <- read_stata(file = "data/diabetes.dta")
```

# Manipulating Data in R

## Manipulating Data: R Objects

"Everything in R is an object" Chambers (2008), but what does this mean...

- Objects are data structures that we can manipulate
- We refer to these data structures by using variables (i.e `diabetes`)
- Objects are created by assigning values, `<-`, to a variable

```{r first-R-object, exercise = TRUE}
# Create a variable called x and assign it the value "Hello World!"
x <- "Hello World!"
x

# Try creating a variable called 'y' and assign it the value 10

# Change the value of x to pi

```

## Manipulating Data: 'Atomic' Data Types

Every R object can be broken down into a set of 'atomic' R objects, sometimes referred to as data types. The three we will primarily work with are:

- numeric (2, 3.14)
- character ("positive", "los angeles")
- logical (TRUE, FALSE)

To determine the type of an R object, use the `class()` function.

```{r data-types, exercise = TRUE} 
# numeric, L is used to indicate that this is an integer
a_double <- 6.2
a_int <- 10L

# character
city <- "los angeles"

# logical (TRUE or FALSE)
is_it_true <- FALSE

# check type of variables
class(a_double)
class(a_int)
class(city)
class(is_it_true)
```

## Manipulating Data: Data Structures

The primary data structures that you will use on an everyday basis:

- Vectors
- Matrices
- Lists
- Data Frames (a special type of list)

Each data structure has a set of functions that we can use to:

- Access elements
- Add elements
- Modify elements
- Delete elements

Our primary focus will be on **Data Frames**

## Manipulating Data: Vectors

To use data frames, we first need to understand vectors

  - Vector: sequence of numbers, characters, **OR** TRUE/FALSE values
  - Most common way to create a vector is the `c()` function
  - `length()` returns how many values are stored in a vector
  - `typeof()` returns the type of data stored in a vector
  
```{r vectors, exercise = TRUE}
# A vector to store the age of everyone
age <- c(25, 30, 47, 38, 29, 72, 22, 55)

# How many age values are stored in age
length(age)

# What type of data is in age
class(age)
```

## Manipulating Data: Vectors (Accessing)

Two primary ways to access data in a vector, both use `[]` to access

1) By providing the position(s) of the elements in the vector
    - `my_subvec <- myvec[c(2, 10, ...)]`
    - Use another vector to specify the position(s)!
    
2) By logical subsetting
    - `my_subvec <- myvec[c(TRUE, FALSE, ...)]`
    - Example: I want all values that are less than 40

**Values by position**

```{r vectors2, exercise = TRUE}
# A vector to store the age of everyone
age <- c(25, 30, 47, 38, 29, 72, 22, 55)

# Get the values stored at positions 3, 4, and 7
subset_by_position <- age[c(3, 4, 7)]
```

**Values by logical subsetting**

```{r vectors3, exercise = TRUE}
# A vector to store the age of everyone
age <- c(25, 30, 47, 38, 29, 72, 22, 55)

# Determine what values are less than 40
is_lt_40 <- age < 40
is_ls_40

# Get values that are less than 40

# Use the TRUE/FALSE vector we created
subset_logical <- age[is_lt_40]

# Compute TRUE/FALSE vector directly in []
subset_logical2 <- age[age < 40]
```

## Manipulating Data: Vectors (Modifying)

The same methods used to access values can be used to change values

```{r vectors4, exercise = TRUE}
# A vector to store the age of everyone
age <- c(25, 30, 47, 38, 29, 72, 22, 55)

# Change the values at positions 3, 4, and 7 to zero
age[c(3, 4, 7)] <- 0

# Get all values that are less than 40 and set to zero
age <- c(25, 30, 47, 38, 29, 72, 22, 55)
age[age < 40] <- 0
```

To add elements to the end (or middle/beginning), use `c()` or `append()`

```{r vectors5, exercise = TRUE}
a_vec <- c(7, 18, 22.2, -4, 0)
more_num <- c(10, 8, 12)

# add more_num to end of a_vec
c(a_vec, more_num)

# add more_num after 2nd element in a_vec
append(a_vec, more_num, after = 2)
```

## Manipulating Data: Vectorization

A number of operations in R are 'vectorized', vectorization tends to lead to

- Cleaner and more succinct code
- A more efficient implementation of the operations 

All the operations below are 'vectorized'

```{r vectorization, exercise = TRUE}
a_vec <- c(7, 18, 22.2, -4, 0)

# add 10 to each element in a_vec
a_vec + 10 

# square each element in a a_vec 
a_vec^2 # square each element

# test whether each element in a_vec is < 0
a_vec < 0

# access all elements with value < 0
a_vec[a_vec < 0] 
```

## Manipulating Data: Data Frames

A `data.frame` is used to store and analyze rectangular data

  - Similar to an Excel sheet: 
    - Rows = Observations 
    - Columns = Variables
    
  - Several R packages have extended the functionality of `data.frame`
    - `tibble`: A `data.frame` with additional features from the Tidyverse
    - `data.table`: Improved efficiency to manipulate large data sets
    
  - All of the functions used to read data so far end up creating data frames!
  - For the rest of this discussion, we will use the `tibble` form produced by `readr`
  
*By definition, data frames are just a list of equal-length vectors*

## Manipulating Data: Data Frames (Structure)

After reading in data, it is important to verify that the data was read correctly

  - Use `head(df)` to print out the first few rows of the data
  - `dim(df)`, `nrow(df)`, and `ncol(df)` to inspect the dimensions
  - `typeof(df)`, `str(df)`, and `class(df)` provide additional info
  - Use `str()` to verify variable type (i.e. numeric or character)

```{r data-frames, exercise = TRUE, exercise.lines=15}
# Print first 5 rows
head(dbts_df)

# Dimensions of the data set
dim(dbts_df)
nrow(dbts_df)
ncol(dbts_df)

# Additional (helpful) information
str(dbts_df)
```

## Manipulating Data: Data Frames (Accessing)

Three primary ways to access subsets of a `data.frame`

1) By providing row number(s) and/or column number(s)
    - `mydf[c(3, 5, ...), c(1, 4, ...)]`
  
2) By providing row names(s) and/or column name(s)
    - `mydf[c("row1", "row2"), "mycol1"]`
    - To access a single column, can use
        - `mydf$column_name`, returns vector
        - `mydf[['column_name']]`, returns vector
        - `mydf['column_name']`, returns data frame
  
3) By logical subsetting
    - `mydf[c(TRUE, FALSE, ...), c(FALSE, TRUE, ...)]`
    - Example: Get all rows where the person is at least 50 years old

- If rows/columns not specified, R returns ALL rows/columns by default
    - `mydf[, c(1, 4)]` returns all rows from columns 1 and 4

## Manipulating Data: Data Frames (Accessing)

**Get data by row/column position**

```{r data-frames2, exercise = TRUE, exercise.lines=5}
# Return the rows 2, 5, and 10 and only the 3rd column
dbts_df[c(2, 5, 10), 3]
```

**Get data by row/column name**

```{r data-frames3, exercise = TRUE, exercise.lines=5}
# Return all rows from the glucose and diabetes variables
dbts_df[, c("glucose", "diabetes")]

# Return only the diabetes variable using $
dbts_df$diabetes
```

**Get data by logical vectors**

```{r data-frames4, exercise = TRUE, exercise.lines=20}
### Return all rows where age > 40 and diabetes is "pos" ###

# Who is over 40
over40 <- dbts_df$age > 40

# Who is diabetic
is_diabetic <- dbts_df$diabetes == "pos"

# Who is over 40 AND diabetic
over40_diabetic <- over40 & is_diabetic

# Get all observations that are over 40 and diabetic
dbts_df[over40_diabetic, ]

# Same results in one line
dbts_df[dbts_df$age > 40 & dbts_df$diabetes == "pos", ]
```

## Logical Operators

- <, >, <=, >=
- == (equal)
- != (not equal)
- || (OR), 
- \| (OR, vectorized, element-wise)
- && (AND), 
- & (AND, vectorized, element-wise)

```{r logical-operators, exercise=TRUE}
# Create two numeric variables, x and y, assign them values
x <- 3
y <- 7
# Create two Boolean variables, t and f, assign them values
t <- TRUE
f <- FALSE
# Compare numeric values using logic statements
y > x
y == x
# Compare &&, ||
t && f
t || f
```

**Compare && and || to & and |**

- && and || return scalar value using only 1st element of each object being compared
- & and | return logical vector

```{r logical-vector, exercise = TRUE}
c(TRUE, FALSE, TRUE) && c(TRUE, TRUE, FALSE)
c(TRUE, FALSE, TRUE) & c(TRUE, TRUE, FALSE)
```

## Manipulating Data: Data Frames (Modifying)

Similar syntax can be used to modify a `data.frame`:

**Add a new variable (column) to a data frame, mydf$new_variable_name**
```{r data-frames5, exercise = TRUE, exercise.lines=5}
# Create variable is_diabetic (TRUE = "pos", FALSE = "neg")
dbts_df$is_diabetic <- dbts_df$diabetes == "pos"
```

**Modify an existing variable in a data frame**
```{r data-frames6, exercise = TRUE, exercise.lines=5}
# Change is_diabetic to numeric (TRUE = 1, FALSE = 0)
dbts_df['is_diabetic'] <- as.numeric(dbts_df$diabetes == "pos")
```

**Remove a variable from a a data frame, set to NULL**
```{r data-frames7, exercise = TRUE, exercise.lines=5}
# Remove is_diabetic from data frame
dbts_df$is_diabetic <- NULL
```

## An aside: dplyr, readr, ..., the Tidyverse

The Tidyverse is a set of R packages intended to make data manipulation and analysis

`readr`, `readxl`, `haven`

- Provide a number of functions to read data from different sources

`dplyr`

- Provides a clean (SQL-like) syntax to shape your data exactly how you need it
- Naturally extends to when your data lives in multiple data sets
- `tidyr` extends functionality to ensure you have 'tidy' data

`tibble`

- An improved `data.frame` that you have already been using through this section



# Summarizing Data in R

## Summarizing Data: Summary Statistics + Plots

**For categorical variables:**

- Summary statistics
    - Frequency counts 
    - Proportions
    
- Plots
    - Bar chart

**For continuous variables:**

- Summary statistics
    - Measures of centrality (mean, median, mode)
    - Spread/variation (variance, standard deviation, range)
    - Quantiles/percentiles (min, max, 25% , 50% (median), etc..)
    
- Plots
    - Histogram
    - Boxplot (+ individual points)

## An Aside: The ggplot2 package

For the plots in this tutorial, we will be using the popular R package `ggplot2`

Base R does contain plot functions, but....

- `ggplot2` allows for quick creation of visually appealing plots (they look nice)
- "Easy" to use interface with lots of customization

Every ggplot starts with `ggplot(data, aes(x=, y=, ...))` to define

- What `data` to use
- Which variables in `data` to use and how they will be used
- ggplot allows for us to add to the plot in 'layers'
    - `ggplot(...) + geom_point(...) + labs(...) + ...`
    - We don't have to specify the details all in one function

## Summarizing Data: Categorical Variables

`table()` is the standard for tabulating counts of each unique value in a variable

- Use extends to cross-tabulating two variables (i.e. contingency table)

`prop.table()` can be used to obtain proportions

```{r cat-desc, exercise = TRUE, exercise.lines=10}
# Find counts of diabetic / non-diabetic
diabetic_counts <- table(dbts_df$diabetes)
diabetic_counts

# Get proportions from counts
diabetic_prop <- prop.table(diabetic_counts)
diabetic_prop
```

## Summarizing Data in R: Categorical Variables

`geom_bar()` is used to create a bar chart for categorical variables

- For a bar chart, we need to tell `geom_bar()` what variable is on the x-axis
  - `aes()` is used to specify this information

```{r cat-plot, exercise = TRUE, exercise.lines=3}
# Create bar chart of diabetes variable
ggplot(data = dbts_df, aes(x = diabetes)) + geom_bar()
```

## Summarizing Data: Continuous Variables

- Measures of centrality
    - `mean()`, `median()`, `mode()`

- Spread / Variation
    - `var()`, `sd()`, `range()`
    
- Quantiles / Percentiles
    - `min()`, `max()`, `quantile()`
    
```{r cont-desc, exercise = TRUE, exercise.lines=15}
# Compute mean, median, mode of glucose
mean(dbts_df$glucose)
median(dbts_df$glucose)
mode(dbts_df$glucose)

# Compute variation of glucose
var(dbts_df$glucose)
sd(dbts_df$glucose)
range(dbts_df$glucose)

# Compute quantiles of glucose
min(dbts_df$glucose)
max(dbts_df$glucose)
quantile(dbts_df$glucose, probs = c(0.25, 0.75))

# Summary function computes multiple statistics
summary(dbts_df$glucose)
```

## Summarizing Data: Continuous Variables

- `geom_histogram()` is used to generate a histogram
- `geom_boxplot()` is used to generate a boxplot
    - Using `y=` creates a vertical boxplot, `x=` creates a horizontal boxplot
    - We can overlay data points on boxplot with `geom_jitter()`
    
```{r cont-plot, exercise = TRUE, exercise.lines=10}
# Histogram of glucose
ggplot(dbts_df, aes(x = glucose)) + geom_histogram(binwidth = 10)

# Boxplot of glucose
ggplot(dbts_df, aes(y = glucose)) + geom_boxplot()

# Boxplot of glucose + data points
ggplot(dbts_df, aes(x = factor(0), y = glucose)) + geom_boxplot() + 
  geom_jitter(shape = 16, position = position_jitter(0.2), alpha = .3)
```

## Summarizing Data: Bivariate Analyses

The next step is to analyze pairwise relationships between variables

**Continuous vs. Continuous:**

- Correlation: Strength of the linear relationship between two variables
- Scatterplot: Bivariate point plot to visualize relationship

**Continuous vs. Categorical:**

- Stratified Boxplot: Boxplot of continuous variable for each level of categorical

**Categorical vs. Categorical:**

- Contingency Table: Cross-tabulate counts for each unique combination of the two categorical variables

## Summarizing Data: Continuous vs. Continuous

The correlation between two variables can be found with the `cor()` function

```{r corr, exercise = TRUE, exercise.lines=3}
cor(dbts_df$glucose, dbts_df$insulin)
```

`corrplot()` from the package R *corrplot* makes it easy to visualize the correlation between all variables 

```{r corr2, exercise = TRUE, exercise.lines=3}
corrplot(cor(dbts_df[, -9]), type = "lower")
```

## Summarizing Data: Continuous vs. Continuous

`geom_point()` creates a scatterplot between two continuous variables

```{r scatter-plot, exercise = TRUE, exercise.lines=3}
ggplot(dbts_df, aes(x = glucose, y = insulin)) + geom_point()
```

A trend line can be added by using `geom_smooth()`

```{r scatter-plot2, exercise = TRUE, exercise.lines=3}
# Scatterplot with Linear Trend
ggplot(dbts_df, aes(x = glucose, y = insulin)) + geom_point() +
  geom_smooth(method = "lm")
```

```{r scatter-plot3, exercise = TRUE, exercise.lines=3}
# Scatterplot with LOESS
ggplot(dbts_df, aes(x = glucose, y = insulin)) + geom_point() +
  geom_smooth(method = "loess")
```

## Summarizing Data: Continuous vs. Categorical

Group-specific means can be computed with the `aggregate()` function:

```{r group-means, exercise = TRUE, exercise.lines=3}
aggregate(dbts_df[, -9], list(dbts_df$diabetes), mean)
```

Same result, different syntax:

```{r group-means2, exercise = TRUE, exercise.lines=3}
aggregate(. ~ diabetes, dbts_df, mean)
```

`geom_boxplot()` can be used to create side-by-side boxplots based on a group:

```{r group-boxplot, exercise = TRUE, exercise.lines=3}
ggplot(dbts_df, aes(y = glucose, x = diabetes, fill = diabetes)) + 
  geom_boxplot()
```

## Summarizing Data: Categorical vs. Categorical

`table()` can be used to create a contingency table

Dichotomize glucose (< 126 vs. >= 126), create table of glucose vs. diabetes

  - `ifelse` is a shorthand to conditionally assign values based on a logical vector
  - `ifelse(c(TRUE, FALSE), "Value if TRUE", "Value if FALSE")`

```{r contingency-table, exercise = TRUE, exercise.lines=8}
# Dichotomize glucose at 126
high_glucose <- ifelse(dbts_df$glucose >= 126, ">= 126", "< 126")

# Contingency table of age vs diabetes
table(high_glucose, dbts_df$diabetes)
```

`prop.table()` can again be used to obtain proportions

  - `margin=` option used to compute percentages along rows or columns
  - 1 = rows, 2 = columns

```{r prop-table, exercise = TRUE, exercise.lines=8}
# Dichotomize glucose at 126
high_glucose <- ifelse(dbts_df$glucose >= 126, ">= 126", "< 126")

# Contingency table of age vs diabetes
prop.table(table(high_glucose, dbts_df$diabetes))
```

Suppose we want to predict if someone has diabetes...how well does `high_glucose` classify people?

```{r glucose-classify, exercise = TRUE, exercise.lines=20}
# Dichotomize glucose at 126
high_glucose <- ifelse(dbts_df$glucose >= 126, ">= 126", "< 126")

# Compute sensitivity and specificity
prop.table(table(high_glucose, dbts_df$diabetes), margin = 2)

# Make sure we are getting the right answer
sensitivity <- 176 / (176 + 92)
specificity <- 379 / (379 + 121)

# Compute positive predictive value and negative predictive value
prop.table(table(high_glucose, dbts_df$diabetes), margin = 1)

# Make sure we are getting the right answer
ppv <- 176 / (176 + 121)
npv <- 379 / (379 + 92)
```

# Hypothesis Testing in R

## Hypothesis Testing: 2-Sample t-test

A two-sample t-test is used to test if there is a mean difference in a continuous outcome between two groups

  - The outcome is assumed to be normally distributed in each group
 
  $$y_1 \sim N(\mu_1, \sigma^2_1)$$
  
 $$y_2 \sim N(\mu_2, \sigma^2_2)$$

  - The null and alternative hypotheses to test for a difference in the means are:
  
  $$H_0: \mu_1=\mu_2$$

 $$H_1: \mu_1 \ne \mu_2 \space (two-sided)$$
 $$H_1: \mu_1 > \mu_2 \space (one-sided)$$
 
$$H_1: \mu_1 < \mu_2 \space (one-sided)$$

  - The test statistic follows the Student's t distribution and is computed as:
  
  $$T=\frac{\bar{x}_1-\bar{x}_2}{s_p\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}} \space (equal \space variance)$$

$$T=\frac{\bar{x}_1-\bar{x}_2}{\sqrt{\frac{s^2_1}{n_1} + \frac{s^2_2}{n_2}}} \space (unequal \space variance)$$

## Hypothesis Testing: 2-Sample t-test

Below is a small simulation that simulates an outcome in two groups:

- Use the inputs to change the mean and standard deviation (SD)
- Results of two-sample t-test are shown below stratified boxplot

```{r, context="render", echo = FALSE}

shinyUI(fluidPage(
  sidebarPanel(
    helpText("Group 1:"),
    numericInput("mu1", label = "Mean", value = 0, step = 0.1),
    numericInput("sd1", label = "SD", value = 1, step = 0.1),
    helpText("Group 2:"),
    numericInput("mu2", label = "Mean", value = 0, step = 0.1),
    numericInput("sd2", label = "SD", value = 1, step = 0.1)
  ),
  mainPanel(
    plotOutput("boxPlot"),
    tableOutput("ttestTable")
  )
))
```

```{r, context="server"}

dat <- reactive({
  n <- 200
  y_group1 <- rnorm(n, input$mu1, input$sd1)
  y_group2 <- rnorm(n, input$mu2, input$sd2)
  dat <- data.frame(
    group = c(rep("Group 1", n), rep("Group 2", n)), 
    y = c(y_group1, y_group2)
  )
})

output$boxPlot <- renderPlot({
    ggplot(dat(), aes(x = group, y = y, fill = group)) + 
      geom_boxplot()
})

output$ttestTable <- renderTable({
    dat_cur <- dat()
    if (input$sd1 == input$sd2) {
      ttest <- t.test(y ~ group, dat_cur, var.equal = TRUE)
    } else {
      ttest <- t.test(y ~ group, dat_cur, var.equal = FALSE)
    }

    results <- data.frame(
      mean_grp1 = mean(dat_cur$y[dat_cur$group == "Group 1"]),
      mean_grp2 = mean(dat_cur$y[dat_cur$group == "Group 2"]),
      sd_grp1 = sd(dat_cur$y[dat_cur$group == "Group 1"]),
      sd_grp2 = sd(dat_cur$y[dat_cur$group == "Group 2"]),
      t_value = ttest$statistic,
      p_value = ttest$p.value
    )
    colnames(results) <- c(
      "Mu1", 
      "Mu2", 
      "SD1", 
      "SD2", 
      "T", 
      "p-value"
    )
    results
}, digits = 4)
```

## Hypothesis Testing: 2-Sample t-test

Returning back to the diabetes data set...

`t.test()` is used to test if mean glucose differs by diabetes status

- There are two ways we can specify:
    1) `t.test(y ~ group, data = dat)`
    2) `t.test(y[grp == "Group 1"], y[grp == "Group 2"])`

```{r ttest-glucose, exercise = TRUE, exercise.lines=12}
# 2-sample t-test for glucose by diabetes status

# Using a formula
t.test(glucose ~ diabetes, data = dbts_df)

# Passing each outcome separately by group
t.test(
  dbts_df$glucose[dbts_df$diabetes == "pos"], 
  dbts_df$glucose[dbts_df$diabetes == "neg"]
)
```

## Hypothesis Testing: 2-Sample t-test

Let's focus on the type of object returned by `t.test()`...a **List**

Lists are 'generic' vectors that can be composed of different R objects.

- A `data.frame` is a List object composed of vectors
  
Investigate the object returned by `t.test()`:

```{r ttest-glucose2, exercise = TRUE, exercise.lines=22}
# 2-sample t-test for glucose by diabetes status
result <- t.test(glucose ~ diabetes, data = dbts_df)

# What is the structure of result
str(result)

# Accessing elements in a list is the same as getting columns in a data.frame

# Using the lst$element_name
result$statistic

# Using lst[['element_name']
result[["statistic"]]

# To return a sublist [c("elem1", "elem2")]
result[c(1, 3)]
result[c("statistic", "p.value")]

# In some cases, it makes sense to return a vector
unlist(result[c("statistic", "p.value")])
```

## Hypothesis Testing: Chi-square Test

A Chi-Square test allows us to test whether there is an association between two categorical variables

- The null and alternative hypotheses can be expressed as:
    - $H_0$: There is no association between the categorical variables
    - $H_1$: There is an association between the categorical variables
    
- The test statistic for the Chi-Square test compares the observed and expected counts:

$$\chi^2=\sum \frac{(O-E)^2}{E}$$

`chisq.test()` is used to compute a Chi-Square test in R

- Below, we look at `high glucose` vs. `diabetes status`
    
```{r chisq-test, exercise = TRUE, exercise.lines=16}
# Chi-Square test for high glucose and diabetes status
high_glucose <- dbts_df$glucose > 126
chisq.test(high_glucose, dbts_df$diabetes)

# Same test, different syntax
chisq_result <- chisq.test(table(high_glucose, dbts_df$diabetes))

# Get observed counts and expected counts
observed <- chisq_result$observed
expected <- chisq_result$expected

# What type of data structures are observed and expected?
class(observed)
class(expected)
```

## Hypothesis Testing: Linear Regression

Linear regression is a common method to model the association of a continuous outcome as a linear function of a set of predictors

The general model is expressed as:

$$Y = \beta_0+\beta_1 X_1+\beta_2X_2+...+\beta_pX_p+\epsilon$$
$$\epsilon \sim N(0, \sigma^2)$$
`lm()` is used to fit linear regression models in R

  - As with `t.test()`, the model can be specified two different ways
  - `lm(y ~ x1 + x2 + ..., data = dat)`
  - `lm(y = y, x = x)`, `x` must be a Matrix
  
Let's show that `lm()` and `t.test()` can lead to the same results:

```{r lm-glucose, exercise = TRUE, exercise.lines=5}
# 2-sample t-test for glucose by diabetes status
result_ttest <- t.test(
  glucose ~ diabetes, 
  data = dbts_df, 
  var.equal = TRUE
)

# Fit linear regression model of glucose on diabetes status
result_lm <- lm(glucose ~ diabetes, data = dbts_df)

# View results of regression with summary() function
summary(result_lm)

# Get test statistic for t-test
teststat_ttest <- result_ttest$statistic

# Get test statistic for lm
teststat_lm <- summary(result_lm)$coefficients[2, 3]

# Are they the same?
all.equal(abs(teststat_ttest), abs(teststat_lm), check.attributes = FALSE)
```

## Hypothesis Testing: Linear Regression

Many statistical procedures have additional functions to help parse results

The object returned by `lm()` can be used with the following functions:

  - `summary()` summarize the results of the model
  - `coef()` to return the coefficients from the model
  - `residuals()` to return the residuals from the model
  - `predict()` to use the model make predictions on a new set of data
  - `plot()` plots to assess model fit and assumptions

```{r lm-glucose2, exercise = TRUE, exercise.lines=35}
# Fit linear regression model of glucose on diabetes status
result_lm <- lm(glucose ~ diabetes, data = dbts_df)

# Model fit plots
plot(result_lm)
```

```{r lm-glucose3, exercise = TRUE, exercise.lines=35}
# Fit linear regression model of glucose on diabetes status
result_lm <- lm(glucose ~ diabetes, data = dbts_df)

# Create our own residuals vs fitted plot

# Get residuals and fitted (predictions)
resids <- residuals(result_lm)
fitted <- predict(result_lm, newdata = dbts_df)

# Create data frame
model_results <- data.frame(
  resids = resids,
  fitted = fitted
)

# Compute mean of residuals at fitted values
mean_results <- aggregate(
  resids ~ fitted, 
  data = model_results, 
  mean
)

ggplot(model_results, aes(x = fitted, y = resids)) + 
  geom_point() +
  geom_line(data = mean_results, col = "red") +
  labs(x = "Fitted", y = "Residuals") +
  ggtitle("Plot of fitted vs. residuals")
```

## Summary

In this tutorial, you learned how to:

  - Read data from various data sources
  - Manipulate data in various froms
  - Compute summary statistics and generate plots to visualize your data
  - Conduct statistical analyses and use the results returned
